---
title: "ESM244lab1"
author: "MaddieFrick"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#attach packages
library(tidyverse)
library(here) #for nice paths to read in data in a project
library(sf) #spacial data wrangling
library(tmap) #to make an interactive map
```

### Read in SF trees data 

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv")) #gives a robust file path
```

### Basic wrangling reminders

Refresh some skills for data wrangling and summary statistics using functions in 'diplyr' package

Find the top five highest observations of trees by legal_Status, do some wrangling, make a graph

```{r}
top_five_status <- sf_trees %>% 
  count(legal_status) %>%  #count recognizes finds counts of observations based on one or more variables :groups (replacing group by), finds counts (replaces n), puts it in a table (replaces summarize)
  drop_na(legal_status) %>% #drops all of the NA row where NA is in the legal status row
  rename(tree_count = n) %>%  #new name left old name right
  relocate(tree_count) %>% #this brings tree count to the first column position
  slice_max(tree_count, n = 5) #allows to identify by highest value in a variable
```

Make a graph of those five observations

```{r}
ggplot(data = top_five_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col() + #reorder thing orders the status by the tree count instead of alphabetical 
  labs(x = "Legal Status", y = "Tree Count")+
  coord_flip() + #flips the x and y axis
  theme_minimal()
 
```
More data wrangling
only want to keep observations for blackwood acacia trees
```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% #checks wherever this string shows up at all
  select(legal_status, date, latitude, longitude)

ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()


```
### tidyr::separate() and unite() functions
useful for combining or separating columns

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::") #break it up into two columns

```

Example: tidyr::unite()
```{r}
sf_tree_unite <- sf_trees %>% 
  unite("id_Status", tree_id:legal_status, sep = "_cool!_")


```

Make some actual maps of blackwood acacia trees in SF!
'st_as_sf()' to convert latitude and longitude to spatial coordinates

```{r}
blackwood_acacia_sp <- blackwood_acacia %>% 
  drop_na(longitude, latitude) %>% 
  st_as_sf(coords = c("longitude", "latitude")) #whatever the variable names are storing the long and lat

st_crs(blackwood_acacia_sp) = 4326 #set it as this four digit code from the coordinate reference system

ggplot(data = blackwood_acacia_sp) +
  geom_sf(color = "darkgreen") #this works because it is spacial data

```
Read in SF roads shapefile 
```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

st_transform(sf_map, 4326)

ggplot(data = sf_map) +
  geom_sf()
```
combine acacia tree observations and the road map
```{r}
ggplot() +
  geom_sf(data = sf_map, size = 0.1, color = "darkgray") +
  geom_sf(data = blackwood_acacia_sp, color = "red", size = 0.5) +
  theme_void()
```
